Testing Traversability of Protein Tunnels with Flexible Ligands Using Motion Planning

### 2017-06-29 paper.tex
images

To guide the growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of sphere of the tunnel.
The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated around the sphere $t_v \in T$.
After the tree reaches the sphere $t_v$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

@@

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::main}Main loop of the RRT planner}
\KwIn{
    tunnel spheres $c_i$ and their radii $r_i$, $i=1,\ldots,n$, 
    initial configuration $\qinit$,
    goal configuration $\qgoal$,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 1$; // index of the virtual goal\\
$iteration = 0$\;
\While{$iteration < \Imax$ {\bf and}  $v < n$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around actual virtual goal $c_n \in T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
    expand($\qnear$)\;
    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $c_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
\end{algorithm}

@@

\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::expand}Expand}
\KwIn{
   configuration $\qnear$ to be expanded,
   random configuration $\qrand$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$, and $\sdelta$
}
%\KwOut{
%    set $S$ of collision-free configurations reachable from $\qnear$\;
%}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
        $\qnew = \emptyset$; // empty configuration\\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$)}{
                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qnear) < \da(\qnew,\qnear)$}{ \nllabel{alg::expand:a}
                    $\qnew = q$\;
                }
            } 
        }
        \If{$\qnew \ne \emptyset$} {
            $\T$.addNode($\qnew$)\;
            $\T$.addEdge($\qnear,\qnew$)\;
            {\bf break;} // go to next conformation
        }
    }
}
%\return $S$\;
\end{algorithm}

The result of each planning trial is the tree $\T$ of collision-free configurations.
The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the nearest node in the tree towards $\qgoal$, is found in the tree.
The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.

@@

The trajectories are computed for all initial configurations $\qinit \in \QI$ and for each of the starting
configuration, multiple trajectories are obtained.
For the purpose of this paper, we compute the tunnel characteristics using all computed trajectories.

Trajectory $P$ reaches the tunnel sphere $c_i \in t$, if the 3D Euclidean distance of the 
nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
Let $N(i)$ denote the number of trajectories that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput, and the scaling factor.

The accessibility $A(i)$ of the sphere $i$ is the probability of reaching the sphere by the computed trajectories,
i.e., $A(i)=N(i) / N$, where $N$ is the total number of trajectories.
The accessibility  shows how probable is to pass the tunnel up to the sphere $i$.
The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
the bottleneck have low accessibility.

@@

%In order to fursupport the tunnel exploration process performed by biochemists.
The computed properties of the tunnel are presented visually by mapping the accessibility, throughput, and scale to the sphere representation
of the tunnel (Fig.~\ref{fig:properties}) using a color map.
Alternatively, the surface representation can be used to visualize the tunnel.
In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $t_i$ is the closest to the point among all tunnel parts.

@@

The mapping using colors helps to identify difficult parts of the tunnel, but it is also necessary to show the original trajectories.
This is necessary for example if the trajectories detour from a part of the tunnel.
The na\"{i}ve visualization of all trajectories is possible, but too slow if we are dealing with thousands of trajectories.

In order to visualize the most typical trajectories, we cluster the trajectories according to their similarity.
The number of configurations in the trajectories usually differ due to the stochastic nature of RRT and it also depends on the
initial configurations.

To compare two trajectories (which are sequences of configurations in the configuration space), we first convert them
to a vector $v=(x_1,x_2,\ldots,x_M)$, where $x_i \in \R^3$ is a 3D
point representing the trajectory and $M$ is the parameter determining how precisely we want to represent the trajectories.
To obtain the vector $v$ for a given trajectory $P$, first a set of spheres 
 $S_1, ..., S_N$, where $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
%First, we represent each trajectory $t$ by a vector $v = (x_1, x_2, ..., x_N)$ of $N$ samples where $x_i \in \mathbb{R}^3$.
%In order to obtain the samples, we find the average starting point $x_s$ of all trajectories and the most distant (w.r.t. $x_s$) ending point $x_e$ also among all trajectories.
%Using $x_s$ and $x_e$ we define spheres $S_1, ..., S_N$ such that $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
Then, a vector $v$ for a trajectory $P$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $P$ which is the nearest one to the surface of $S_i$.
Finally, we define a metric $d$ for trajectories $P_i$ and $P_j$ as $d(P_i, P_j) = \frac{1}{N} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
The clustering is then computed using the UPGMA clustering algorithm~\cite{sokal1958statistical}.
When visualizing, we choose a representative from each cluster and depict its trajectory in 3D using a polyline.
Moreover, we map the representative's cluster size onto the width of the polyline.
In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).

@@

>>>>

To guide the growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of sphere of the tunnel.
The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated around the sphere $c_v \in T$.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

@@

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::main}Main loop of the RRT planner}
\KwIn{
    tunnel spheres $c_i$ and their radii $r_i$, $i=1,\ldots,n$, 
    initial configuration $\qinit$
%    goal configuration $\qgoal$,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 1$; // index of the virtual goal\\
$iteration = 0$\;
\While{$iteration < \Imax$ {\bf and}  $v < n$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around actual virtual goal $c_n \in T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
    expand($\qnear$)\;
    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $c_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
\end{algorithm}

@@

\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::expand}expand}
\KwIn{
   configuration $\qnear$ to be expanded,
   random configuration $\qrand$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$, and $\sdelta$
}
%\KwOut{
%    set $S$ of collision-free configurations reachable from $\qnear$\;
%}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
        $\qnew = \emptyset$; // empty configuration\\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$)}{
                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qnear) < \da(\qnew,\qnear)$}{ \nllabel{alg::expand:a}
                    $\qnew = q$\;
                }
            } 
        }
        \If{$\qnew \ne \emptyset$} {
            $\T$.addNode($\qnew$)\;
            $\T$.addEdge($\qnear,\qnew$)\;
            {\bf break;} // go to next conformation
        }
    }
}
\end{algorithm}

The result of each planning trial is the tree $\T$ of collision-free configurations.
The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the nearest node in the tree towards $\qgoal$, is found in the tree.
The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
The scale of ligand may vary in each part of the trajectory and also the conformations.
The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.

@@

For each initial configuration $\qinit \in \QI$, $K$ trajectories are computed, which results in the set of $K |\QI|$ trajectories.
All these trajectories are used to compute the following properties of the tunnel $T$.
A trajectory $P$ reaches the tunnel sphere $c_i \in T$, if the 3D Euclidean distance of the 
nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
Let $N(i)$ denote the number of trajectories that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput, and the scaling factor.

The accessibility $A(i)$ of the sphere $i$ is the probability of reaching the sphere by the computed trajectories,
i.e., $A(i)=N(i) / N$, where $N=K|\QI|$ is the total number of trajectories.
The accessibility  shows how probable is to pass the tunnel up to the sphere $i$.
The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
the bottleneck have low accessibility.

@@

%In order to fursupport the tunnel exploration process performed by biochemists.
The computed properties of the tunnel are presented visually by mapping the accessibility, throughput, and scale to the sphere representation
of the tunnel using a color map.
Alternatively, the surface representation can be used to visualize the tunnel.
In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $t_i$ is the closest to the point among all tunnel parts.
Examples of collow mapping are shown in Fig.~\ref{fig:properties}.

@@

Besides the color mapping, it is also necessary to show the computed trajectories.
This is useful for example to find out in which part of the tunnel the trajectories detour.
The na\"{i}ve visualization of all trajectories is possible, but too slow if we are dealing with thousands of trajectories.

In order to visualize the most typical trajectories, we cluster the trajectories according to their similarity.
The number of configurations in the trajectories usually differ due to the stochastic nature of RRT and it also depends on the
initial configurations.
% by VV ==================================================================
%For the purpose of clusting, the trajectories have to be first represented by vectors of same length.
%\def\cstart{\overline{c_{start}}}
%\def\cend{\overline{c_{end}}}
%Let $\cstart \in \R^3$ and $\cend \in \R^3$ denote the average starting and ending point of all computed trajectories.
%A line $x(t) = \cstart + t(\cend - \cstart), 0 \le t \le 1$ is created. For each point on the line, a radius
%$r(t)$ is assigned as $r(t) =

To compare two trajectories  (which are sequences of configurations in the configuration space), they are first converted
to a vector $v=(x_1,x_2,\ldots,x_M)$, where $x_i \in \R^3$ is a 3D
point representing the trajectory and $M$ is the parameter determining how precisely we want to represent the trajectories.

%by ADAM ============================================================
%To obtain the vector $v$ for a given trajectory $P$, first a set of spheres 
 %$S_1, ..., S_M$, where $S_i = (\cstart, r_i)$ where $r_i = i \frac{|\cstart \cend|}{M + 1}$.
 %First, we represent each trajectory $t$ by a vector $v = (x_1, x_2, ..., x_N)$ of $N$ samples where $x_i \in \mathbb{R}^3$.
%In order to obtain the samples, we find the average starting point $x_s$ of all trajectories and the most distant (w.r.t. $x_s$) ending point $x_e$ also among all trajectories.
%Using $x_s$ and $x_e$ we define spheres $S_1, ..., S_N$ such that $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
Then, a vector $v$ for a trajectory $P$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $P$ which is the nearest one to the surface of $S_i$.

Finally, we define a metric $d$ for trajectories $P_i$ and $P_j$ as $d(P_i, P_j) = \frac{1}{N} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
The clustering is then computed using the UPGMA clustering algorithm~\cite{sokal1958statistical}.
When visualizing, we choose a representative from each cluster and depict its trajectory in 3D using a polyline.
Moreover, we map the representative's cluster size onto the width of the polyline.
In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).

@@

\begin{figure}
\centering
\begin{tabular}{ccc}
\includegraphics[width=0.32\textwidth]{fig/tunne4ac}& 
\includegraphics[width=0.32\textwidth]{fig/tunne4bc} & 
\includegraphics[width=0.32\textwidth]{fig/tunne4cc} \\
(a) & (b) & (c)                        
\end{tabular}
\caption{\label{fig::t4}
    todo
}
\end{figure}
